---
title: "Linear models with categorical explanatory variables"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(mosaic)
crop_growth <- read.csv("www/crop_growth.csv")
knitr::opts_chunk$set(echo = FALSE)
```


## Introduction

In the previous example we looked at linear models with a continuous explanatory
variable, that is:

$$growth = tannin + \epsilon$$
where you will recall that

* growth = growth of caterpillers
* tannin = amount of tannin in the diet
* $\epsilon$ (Greek letter epsilon) = *Error* or *Residual* which is due to 
unknown noise in our experimental data.

Often your single explanatory variable will be categorical, with two or more
different categories, representing different treatments. Examples might include:

| Response | Explanatory variable | Levels |
|:--------:|:--------------------:|:------:|
| Crop growth | Fertiliser | None (control), Nitrogen, Phosphorous, Potassium |
| Biodiversity | Habitat | Woodland, Grassland, Scrub, Suburban |
| qPCR amplification | Treatment | no DNA (-ve control), DNA template (+ve control),  DNA sample A, DNA sample B |

Designed experiments, for example field trials or laboratory experiments, will
often have a *Control* level, i.e. no treatment, whereas ecological surveys may
not have a formal control. Note also that in some laboratory experiments you, such
as the last example, you may have *negative controls*. In the last example you
are wanting to extract the DNA of two samples (A and B), but you have two
controls. The negative control is to check that your buffers etc. are not
contaminated with DNA etc. and whilst your positive control uses a standard
DNA sample where you know what you should expect to check the qPCR amplification
is working properly.

**Note:** 
* When the explanatory variable is cateogrical it is often called a **factor** 
and so you may see the term **factorial analysis** used in some books
* The different categories in your explanatory variable are usually referred
to as **levels**


```{r background_check, echo=FALSE}
question("Which one statement of the following is true?", type="learnr_checkbox",
    answer("If you have categorical variables you must have a control",
           message = "Not always, especially in field surveys"),
    answer("The different categories in a factor are also known as 'levels'",
           message = "Good. You may originally think of them as 'treatments' but
           it is more accurate to call them 'levels", correct = TRUE),
    answer("You can never have a control in an ecological survey",
           message = "With careful planning it is sometimes possible. e.g. if
           you are studying the effects of rabbit grazing on botany, you might
           find a site where no rabbits are present, or fencing is used"),
    answer("Categorical explanatory variables are also known as 'factors'",
           message = "Good. For example, if you have two categorical explanatory variables you
           may see it called '2-factor analysis'", correct=TRUE),
    answer("If you do not apply a treatment it is always called a negative control",
           message = "It is typically referred to simply as a 'control' unless
           you have both positive and negative controls"),
    answer("Negative controls help you check the robustness of your experimental procedures",
           message = "Good. Using both +ve and -ve controls helps you check your procedures
           are running as expected", correct = TRUE),
    allow_retry = FALSE
    )
```


## Example dataset

The data come from an experiment into crop yields on three different soil types,
sand, clay and loam, and as such this is an example where we have a categorical
explanatory variable, with three levels (one for each of the three soil types),
but no formal control. In the R session below, find out the name of the R object,
look at the first 10 rows, check how many replicates there are for each of the
soil types, and calculate the standard deviations for each soil type:

```{r quick_check, exercise = TRUE, exercise.completion = FALSE}

```
<div id="quick_check-hint">
**Hints:**

* The `ls()` function can be used to find out the names of objects in R workspace
* Simply type the name of an object to see all of it, or `head()` for the first part
* Use `summary()` function to get the mean, min, max etc. of continuous variables for
all the data. To break it down nicely per soil type use `favstats(response ~ 
explanatory, data=datasetname)`
* The function `sd(response ~ explanatory, data=datasetname)` and
`tally(~explanatory, data=datasetname)` will display only standard errors and
numbers of replicates respectively, for each soil type.
</div>

You can see that this is a **balanced** experiment in that there are the same
number of replicates for each soil type. Sometimes, in both laboratory and field
studies, data go "missing" for some replicates, but not others. This creates 
**unbalanced** data which can be more complex to analyse.

**Important:** When you ran the `summary()` function you may have noticed that
it returned the minmum, 1st quartile, median, mean, 3rd quartile and maximum of
the `yield` column. It did not do this for the `soil` column, as this is categorical.
However, sometimes your categories might be coded as numbers rather than letters,
and this will confuse R into misreading them as a continuous explanatory. You can
correct this manually, but in general if you have categorical variables, always
encode them with alphanumeric rather than purely numeric.

```{r basic_stats, echo = FALSE}
quiz(caption = "Stats summaries",
  question("The values 14.3, 9.9 and 11.5 are the means for:", type="learnr_radio",
           answer("clay, sand, loam"),
           answer("sand, clay, loam"),
           answer("clay, loam, sand"),
           answer("sand, loam, clay"),
           answer("loam, sand, clay", correct = TRUE),
           random_answer_order = TRUE
           ),
  question("The number of replicates per soil type is:", type="learnr_radio",
           answer("3, 3, 3"),
           answer("9, 9, 9"),
           answer("6, 6, 6"),
           answer("10, 10, 10", correct = TRUE),
           answer("30, 30, 30"),
           random_answer_order = TRUE
           )
)
```

## Plotting your data

Before analysing your data it is always useful to visualise them. One of the
most useful ways of summarising your data when you have a categorical explanatory
variable is via a *boxplot*. You can easily create one of these using the 
`gf_boxplot()` function, which has the syntax:

`gf_boxplot(response_variable ~ explanatory_variable, data=datasetname)`

Try it now on your crop yield data

```{r boxplot, exercise = TRUE, exercise.completion = FALSE}

```
<div id="boxplot-hint">
**Hints:**
* Make sure that you put the response variable on the left and the explanatory
on the right of your equation
* Use the `ls()` function if you've forgotten the name of your R dataset you
want to plot. Use the `head()` function to look at the column names and first
few rows of data.
</div>

### How to interpret a boxplot
A boxplot is made up of several parts. The horizontal lines represent the
**median** values for each of the three soil types (compare it with your data
summaries earlier). The lower and upper parts of the box indicate the lower
75th percentile and the upper 25th percentile. This is a complicated way of
saying that **50% of your data occur within the range of the box**.  This is
also known as the **interquartile range** or IQR. You can calculate the IQR via

`iqr(response_variable ~ explanatory_variable, data = datasetname`

The _**whiskers**_ are 1.5x the IQR. This basically indicates that virtually all
your data are encompassed by the whiskers, which stretch (usually) from the 
minimum to maximum values. However, occasionally there may be **outliers** in
your data, which are indicated by one or more points above or below the whiskers.
You can see that in your data there are outliers for clay and sand.

### What to look for in a boxplot
The main thing to check is that the spread of the data is **_roughly_** the same
for each level of your explanatory data, i.e. that the boxes and whiskers are 
fairly similar. You can see that in your data, they are all approximately the
same size, even though the yield overall appears lower for sand. One assumption
of a linear model is that the spread of values is roughly the same at each level.

### Variations on a boxplot
Two variations on boxplots have become popular in recent years:

* Notch plots. You can create these with

`gf_boxplot(response_variable ~ explanatory_variable, data = datsetname, notch = TRUE)`

* Violin plots, created with

`gf_violin(response_variable ~ explanatory_variable, data = datasetname)`

Experiment with these alternatives in the R console above, and you will be
familiar with them should you see them in any books, papers or lectures.

## Analyse and interpret linear model
Analysis of your data has the same syntax when the explanatory variable is
categorical as when it is continuous, which you have already done. In the box
below, analyse your data with the `lm` command, and summarise results using the
`anova` command.

```{r basic_lm, exercise = TRUE, exercise.completion = FALSE}

```
```{r basic_lm-solution}
# Create the linear model
crop_growth_lm <- lm(yield ~ soil, data = crop_growth)
anova(crop_growth_lm)
```

You can see that against your explanatory variable of soil, there is an F-value
of 4.2447 with 2 and 27 degrees of freedom, which is significant with a p-value
of 0.02495. When writing up your experiment you would report this as

* _"Soil type had a significant effect on crop yield (F<sub>2,27</sub>
= 4.25, p = 0.025)"_

Note how the F-value and p-value were both rounded to 3 signficant figures.

Recall that the F-value gives an indication of the overall importance of an
explanatory variable; the larger the F-value the more important.

## What about the individual levels?
Of course you are probably thinking "it's all very well knowing that the yield
is affected by soil type, but I want to know about whether sand is different
from loam etc.". There are two ways of doing this, one hard, one easy. We'll
look at the harder one first. Re-create your linear model, but this time use
`summary` to see the output:

```{r lm_with_summary, exercise = TRUE, exercise.completion = FALSE}

```
```{r lm_with_summary-solution}
# Create the linear model
crop_growth_lm <- lm(yield ~ soil, data = crop_growth)
summary(crop_growth_lm)
```

This table at looks very similar to the one from the `growth ~ tannin` example
for the caterpillers dataset, where we had a continuous explanatory variable.
What is a bit confusing are the labels for the rows:

* `(Intercept)   11.500`
* `soilloam       2.800`
* `soilsand      -1.600`

**Where has clay gone???**
Look at the numbers again, and the first one might seem familiar from earlier.
Check the means of the three soil types to see if you can work out what is 
going on, with `mean(yield ~ soil, data = crop_growth)`

The row labelled `(Intercept)` is actually the mean value for clay soils. The 
other values are the differences from clay, so:

* 11.500 + 2.800 = 14.3 = loam
* 11.500 - 1.600 =  9.9 = sand

**Why did it pick clay as the intercept?? **
By default, R uses the level with the earliest letter of the alphabet as its
"baseline" `(Intercept)` value to show as an overall mean. You can change this
if you want to using the `relevel()` function, but it is rarely needed.

### How to compare clay, loam and sand easily
Whilst you can use the summary table to derive the information, it is not that
easy to do so (some additional calculations are needed). A much simpler solution
is to use what is amusingly known as the _Tukey Honest Significant Difference_
test, or _Tukey HSD_ test. (Please don't ask me if Tukey also had a Dishonest
Significant Difference test!!). This is available in R using the `TukeyHSD()`
function which is called for you below:

```{r tukey, exercise = TRUE, exercise.completion = FALSE}
crop_growth_lm <- lm(yield ~ soil, data = crop_growth)
TukeyHSD(crop_growth_lm)
plot(TukeyHSD(crop_growth_lm))
```

We have printed out both the table and the plot. The table shows the difference
in yield for each soil type (2.8 for loam versus clay for example), followed by
the 95% confidence intervals for that difference, and finally the p-value. For
the first two comparisons, loam-clay and sand-clay, the 95% confidence intervals
straddle zero, in other words the differnce in yield might not be important.
However, for the sand-loam comparison, on the sand the yield is much lower than
the loam (-4.4: compare with your boxplot). The 95% confidence intervals do not
include zero, and so this difference is signficant with p=0.020

This is obvious in the plot where you can see that it is only for the sand vs 
loam difference that the error bars do not include zero.

## Check model assumptions
After looking at the results of your linear model, it is worth checking that
the model assumptions are valid. Recall that linear models are of the format

$$response\textrm{ }variable = explanatory\textrm{ }variable(s) + \epsilon$$
where you will recall that the Greek letter $\epsilon$ (epsilon) represents the
unknown variability, noise or residual error in your model. This noise has to be
similar for all three soil types in your explanatory variable, and crucially, 
should be normally (bell-shaped) if plotted in a frequency distribution.

We can easily plot the frequency histogram using the `gf_histogram()` function.
Go ahead and try it now. Remember that before you use the `gf_histogram()` function
you will need to use the `residuals()` function to save the positive and 
negative residuals from the model.

```{r resid_histogram, exercise=TRUE, code.completion=FALSE}
crop_growth_lm <- lm(yield ~ soil, data = crop_growth)

```
```{r resid_histogram-solution}
crop_growth_lm <- lm(yield ~ soil, data = crop_growth)
crop_growth_error <- residuals(crop_growth_lm)
gf_histogram( ~ crop_growth_error)
```

As this dataset has 30 records (10 per soil type) it is easier to see that the
frequency histogram approximates to a bell-shaped curve than with the tannins
and growth example that only had 9 records. However, it is still best to check
the QQ (quantile-quantile) plot of your residuals:

```{r resid_qq, exercise=TRUE, code.completion=FALSE}
crop_growth_lm <- lm(yield ~ soil, data = crop_growth)

```
```{r resid_qq-hint-1}
crop_growth_lm <- lm(yield ~ soil, data = crop_growth)
# remember to store your errors in an R object
crop_growth_error <- residuals(crop_growth_lm)
```
```{r resid_qq-hint-2}
crop_growth_lm <- lm(yield ~ soil, data = crop_growth)
crop_growth_error <- residuals(crop_growth_lm)
# Use gf_qq to display the residuals ordered from lowest to highest
gf_qq(~ crop_growth_error)
```
```{r resid_qq-hint-3}
crop_growth_lm <- lm(yield ~ soil, data = crop_growth)
crop_growth_error <- residuals(crop_growth_lm)
gf_qq(~ crop_growth_error) %>% 
  gf_qqline() # Add the theoretical optimum line
```

Whilst there are a couple of outlier residuals (see your original boxplot of
these data), it is clear that most of your residuals roughly follow a straight
line so you can assume that the linear model is valid.